---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-12-01"
editor_options:
  markdown:
    wrap: 72
---

1.) (Use R) Consider the dataset “Homework 6 data.xlsx.” It consists of
5 randomly selected student’s scores on Test 1 and Test 2 in my
introductory statistics course. We want to answer 2 questions:

```{r}
setwd("~/Desktop/Personal_save/Stat_405_Module_14/Module_14_Homework")
#setwd("C:/Users/jake pc/Desktop/Personal_save/Stat_405_Module_14/Module_14_Homework")
HW_6 <- read.csv(file="Homework_6.csv",header=TRUE)
HW_6
```

a.  First, we want to see if there is a difference in the two tests.
    Paired two-tailed t-test

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE)
```

The p-value is less than 0.05, reject the null hypothesis that the means
of Test 1 and Test 2 are equal.

b.  Second, we want to see if there was improvement over the course of
    the semester. H0: Test1 - Test2 \< 0

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE, alternative = "less")
```

Reject the null hypothesis that the mean difference of Test 1 minus Test
2 is equal to zero. Tentatively conclude that the mean difference of
test 1 minus test 2 is less than zero, and therefore that the the grades
of the second test were greater (better) than the first.

2.) (Use R) The data called “plasma” from Anderson et al. (1981)
consists of measurements of plasma concentrations in micromoles/liter
from 10 subjects at times of 8 am, 11am, 2pm, 5 pm, and 8 pm. Analyze
the data in a 1-way ANOVA model choosing time as factor.

```{r}
plasma <- read.csv(file="plasma.csv",header=TRUE)
plasma$time <- factor(plasma$time,levels=c("8am", "11am", "2pm", "5pm", "8pm"), 
                      labels = c("8am", "11am", "2pm", "5pm", "8pm"))

plasma
plasma_model <- lm(plasma ~ time, data = plasma)
anova(plasma_model)
```

The p-value from this test is greater than 0.05, we therefore fail to
reject the null hypothesis that blood plasma levels are not different at
different times.

3.) Two friends play a computer game and each of them repeats the same
level 10 times. The scores obtained are:

```{r}
scores <- read.table(file="scores.txt",header=TRUE)

library(tidyverse)
long <- scores %>%
            pivot_longer(cols = X1:X10, names_to = "trials", values_to = "scores") %>%
                select(-trials)
long$ID <- factor(long$ID, levels = c("Player1","Player2"), labels = c("Player1","Player2"))

write.csv(long, file="long.csv")
```

a.  Player 2 insists that he is the better player and suggests to
    compare their mean performance. Use a t-test to test whether there
    is a difference between their mean performance (alpha = 0.05).

    We are testing for difference of mean on two separate individuals
    —\> 2 sample - unpaired - two sided t-test

```{r}
scores <- t(as.matrix(scores))
colnames(scores) <- scores[1,]
scores <- as_tibble(scores[-1,])
library(dplyr)

scores <- scores %>%
  mutate(across(everything(), as.numeric))

scores



shapiro.test(scores$Player1)
shapiro.test(scores$Player2)


t.test(long$scores ~ long$ID)
```

reject the null hypothesis that the players 2 scores are normally
distributed, therefore player 1 and 2 could never have equal variances

test for difference of scores --\> two-tailed

test results in a p-value of 0.9738, fail to reject the null hypothesis
that the difference in means is equal to zero.

b\. Player 1 insists that he is the better player. He proposes to use
the Wilcoxon rank-sum test for the comparison. What are the results
(alpha = 0.05)?

```{r}
wilcox.test(long$scores ~ long$ID, alternative = "greater")
```

The resulting p-value is less than 0.05, we therefore reject the null
hypothesis that the true location shift is equal to zero and conclude
that player 1 is better than player 2.

4.) (Use R)\
A random sample of 90 adults is classified according to gender and the
number of hours of television watched\
during a week:

\
Use a 0.01 level of significance and test the hypothesis that the time
spent watching television is independent\
of whether the viewer is male or female.

```{r}
table <- matrix(data=c(15,29,27,19),nrow=2,ncol=2,byrow=TRUE,dimnames = list(c("Over 25 hours", "Under 25 hours"),c("Male", "Female")))
table <- t(table)
table
chisq.test(table)
```

The p-value obtained is 0.03338 which is greater than 0.01, we therefore
fail to reject the null hypothesis that time spent watching television
is independent of whether the viewer is male or female.

5.) (Use R)

\
The data set named “Movies” contains a random sample of 35 movies
released in 2008. This sample was\
collected from the Internet Movie Database (IMDb). The goal of this
problem is to explore if the information\
available soon after a movie’s theatrical release can successfully
predict total revenue. All dollar amounts (i.e.,\
variables Budget, Opening, and USRevenue) are measured in millions of
dollars. Consider three explanatory\
variables:

-   The movie’s budget (variable named Budget).

-   Opening weekend revenue (variable named Opening).

-   Number of theaters showing the movie (variable named Theaters).

```{r}
Movies <- read.csv(file="Movies.csv",header=TRUE)
Movies
```

This problem considers using each of these explanatory variables to
attempt to predict a movie’s total US\
revenue (variable named USRevenue).

\
a. Investigate the relationship between the explanatory variable Budget
and response variable USRevenue\
by doing the following:

\
i) Make a scatterplot.

```{r}
plot(x = Movies$Budget, y = Movies$USRevenue)
```

\
ii) Calculate the correlation coefficient.

```{r}
cor(x = Movies$Budget, y = Movies$USRevenue)

```

\
iii) Interpret the scatterplot and correlation coefficient in terms of
trend, strength, and shape.

The correlation coefficient is in the moderate strength range but just
below the strong threshold of plus/minus 0.8, the trend is positive such
that as budget increases US revenue increases. The scatter plot has a
wedge shape such that as budget increases the USRevenue becomes more
variable, this is a problem and will produce a "wedge shape" in the
residuals plot.

\
b. Repeat part (a) for the explanatory variable Opening.

```{r}
plot(x = Movies$Opening, y = Movies$USRevenue)
cor(x = Movies$Opening, y = Movies$USRevenue)
```

The trend is positive such that as the variable opening increase so the
variable USRevenue. The variables have a correlation coefficient of
0.9840177 which is classified as a strong correlation but it is almost
perfect. The relationship between these variables has pattern other than
a linear relationship between predictor and response, and has no
disturbing patters of increasing variability as the predictor variable
increases as the last one did.

\
c. Repeat part (a) for the explanatory variable Theaters.

```{r}
plot(x = Movies$Theaters, y = Movies$USRevenue)
cor(x = Movies$Theaters, y = Movies$USRevenue)
```

The trend is positive such that as variable theaters increases so does
the variable USRevenue. The variables have a correlation coefficient of
0.7153432 which is classified as a moderate correlation. The
relationship between these variables has a very obvious curvature which
appears almost exponential.

\
d. Based on your findings in parts (a) through (c), which of the three
explanatory variables would be most\
appropriate for predicting the response variable USRevenue? Justify your
choice in a few sentences.

Opening is the best choice as predictor variable. The trend is positive
such that as the variable opening increase so the variable USRevenue.
The relationship between Opening and Us Revenue has a positive
correlation coefficient that is almost perfect (.98) and the scatter
plot indicates no pattern other than a positive linearlity. The variable
Budget is completely unsuitable as the residuals of the fitted values
would increase as budget increased. Theaters is unsuitable as it's
relationship is non-linear.

\
e. For the “most appropriate” variable identified in part (d), run a
Simple Linear Regression analysis.

```{r}
Opening_Model <- lm(USRevenue ~ Opening, data=Movies)
plot(USRevenue ~ Opening, data=Movies)
intercept_slope <- coefficients(Opening_Model)
abline(a=intercept_slope[1],b=intercept_slope[2])
summary(Opening_Model)
```

\
f. State the regression equation.

Predicted US Revenue = 3.277\*Opening - 6.9619

\
g. Interpret the slope of the regression line (in context of this data
set).

As Opening sales increases by 1 million, Predicted US Revenue will
increase by 3.277 million.

\
h. Is it meaningful to interpret the y-intercept? Why or why not?

The y-intercept in this model is not meaningful as it's p-value from
testing the null hypothesis that the intercept is equal to zero is
greater than 0.05 and thus the null hypothesis that the true y-intercept
is zero fails to be rejected.

\
i. State r-squared (i.e., the coefficient of determination) and explain
what this value means (in context of\
the data set).

The R-squared of the model is .9683 which means that about 97% of the
variance of the response variable US Revenue is explained by the value
of the Predictor variable opening.

\
j. Use the regression equation from part (f) to predict the total US
revenue for the movie named Get\
Smart. (Budget was 80 million dollars; it was shown in 3911 theaters;
and its opening weekend revenue\
was 38.7 million dollars.) State your predicted value in a sentence that
is in context of the data. Don’t\
forget units!

```{r}
predicted_USRevenue <- predict(Opening_Model, newdata = data.frame(Opening = 38.7))
cat("The linear model predictes that the movie Get Smart which had an opening weekend revenue of 38.7 million dollars would have a total US Revenue of", predicted_USRevenue,"million dollars.")
```

The linear model predicts that the movie Get Smart which had an opening
weekend revenue of 38.7 million dollars would have a total US Revenue of
119.884 million dollars.
