---
title: "Untitled"
output: html_document
date: "2024-12-01"
editor_options: 
  markdown: 
    wrap: 72
---

1.) (Use R) Consider the dataset “Homework 6 data.xlsx.” It consists of
5 randomly selected student’s scores on Test 1 and Test 2 in my
introductory statistics course. We want to answer 2 questions:

```{r}
setwd("~/Desktop/Personal_save/Stat_405_Module_14/Module_14_Homework")
HW_6 <- read.csv(file="Homework_6.csv",header=TRUE)
HW_6
```

a.  First, we want to see if there is a difference in the two tests.
    Paired two-tailed t-test

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE)
```

the p-value is less than 0.05, reject the null hypothesis that the means
of Test 1 and Test 2 are equal.

b.  Second, we want to see if there was improvement over the course of
    the semester. H0: Test1 - Test2 \< 0

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE, alternative = "less")
```

Reject the null hypothesis that the mean difference of Test 1 minus Test
2 is equal to zero. Tentatively conclude that the mean difference of
test 1 minus test 2 is less than zero, and therefore that the the grades
of the second test were better than the first.

**CHECK THIS ONE LATER (Q2)**

2.) (Use R) The data called “plasma” from Anderson et al. (1981)
consists of measurements of plasma concentrations in micromoles/liter
from 10 subjects at times of 8 am, 11am, 2pm, 5 pm, and 8 pm. Analyze
the data in a 1-way ANOVA model choosing time as factor.

```{r}
plasma <- read.csv(file="plasma.csv",header=TRUE)
plasma$time <- factor(plasma$time,levels=c("8am", "11am", "2pm", "5pm", "8pm"), 
                      labels = c("8am", "11am", "2pm", "5pm", "8pm"))

plasma_model <- lm(plasma ~ time, data = plasma)
anova(plasma_model)
```

3.) Two friends play a computer game and each of them repeats the same
level 10 times. The scores obtained are:

```{r}
scores <- read.table(file="scores.txt",header=TRUE)
scores
```

a.  Player 2 insists that he is the better player and suggests to
    compare their mean performance. Use a t-test to test whether there
    is a difference between their mean performance (alpha = 0.05).

    un-paired, two-tailed, t-test

```{r}
transposed_df <- as.data.frame(t(scores), stringsAsFactors = FALSE)
colnames(transposed_df) <- transposed_df[1, ]
transposed_df <- transposed_df[-1, ]
transposed_df$ID <- rownames(transposed_df)
row.names(transposed_df) <- NULL
transposed_df

library(reshape2)
long_scores <- melt(data = transposed_df, id.vars = c("ID"),
                   measured.vars = c("Player1", "Player2"),
                   variable.name = "Player", value.name = "Score")
long_scores$Score <- as.numeric(long_scores$Score)

#you have to check for equality of variances before the t-test because it's a function call option not done automatically 
var.test(long_scores$Score ~ long_scores$Player)
#reject null hypothesis that the variances are equal 
t.test(long_scores$Score ~ long_scores$Player, var.equal = FALSE)

```

P-value is greater than 0.05, fail to reject the null-hypothesis that
the mean score for both players is equal.

b\. Player 1 insists that he is the better player. He proposes to use
the Wilcoxon rank-sum test for the comparison. What are the results
(alpha = 0.05)?

```{r}
wilcox.test(long_scores$Score ~ long_scores$Player)
```

The p-value obtained is 0.03749 which is less than 0.05, we therefore
reject the null hypothesis that the true location shift is equal to
zero.

4.) (Use R)

\
A random sample of 90 adults is classified according to gender and the
number of hours of television watched\
during a week:\

\
Use a 0.01 level of significance and test the hypothesis that the time
spent watching television is independent\
of whether the viewer is male or female.

```{r}
table <- matrix(data=c(15,29,27,19),nrow=2,ncol=2,byrow=TRUE,dimnames = list(c("Over 25 hours", "Under 25 hours"),c("Male", "Female")))
table <- t(table)
table
chisq.test(table)


```

The p-value obtained is 0.03338 which is greater than 0.01, we therefore
fail to reject the null hypothesis that time spent watching television
is independent of whether the viewer is male or female.

\
Note: chisq.test() performs chi-squared contingency table tests and
goodness-of-fit tests in R. By default,\
the continuity correction is used when computing the test statistic for
2 by 2 tables. To use the chi-square test\
without continuity correction, you should add the correct =FALSE option.
Please read the R documentation\
in chisq.test for details.

5.) (Use R)

\
The data set named “Movies” contains a random sample of 35 movies
released in 2008. This sample was\
collected from the Internet Movie Database (IMDb). The goal of this
problem is to explore if the information\
available soon after a movie’s theatrical release can successfully
predict total revenue. All dollar amounts (i.e.,\
variables Budget, Opening, and USRevenue) are measured in millions of
dollars. Consider three explanatory\
variables:

-   The movie’s budget (variable named Budget).

-   Opening weekend revenue (variable named Opening).

-   Number of theaters showing the movie (variable named Theaters).

```{r}
Movies <- read.csv(file="Movies.csv",header=TRUE)
```

This problem considers using each of these explanatory variables to
attempt to predict a movie’s total US\
revenue (variable named USRevenue).

\
a. Investigate the relationship between the explanatory variable Budget
and response variable USRevenue\
by doing the following:

\
i) Make a scatterplot.

\
ii) Calculate the correlation coefficient.

\
iii) Interpret the scatterplot and correlation coefficient in terms of
trend, strength, and shape.

\
b. Repeat part (a) for the explanatory variable Opening.\
c. Repeat part (a) for the explanatory variable Theaters.\
d. Based on your findings in parts (a) through (c), which of the three
explanatory variables would be most\
appropriate for predicting the response variable USRevenue? Justify your
choice in a few sentences.\
e. For the “most appropriate” variable identified in part (d), run a
Simple Linear Regression analysis.\
f. State the regression equation.\
g. Interpret the slope of the regression line (in context of this data
set).\
h. Is it meaningful to interpret the y-intercept? Why or why not?\
i. State r-squared (i.e., the coefficient of determination) and explain
what this value means (in context of\
the data set).\
j. Use the regression equation from part (f) to predict the total US
revenue for the movie named Get\
Smart. (Budget was 80 million dollars; it was shown in 3911 theaters;
and its opening weekend revenue\
was 38.7 million dollars.) State your predicted value in a sentence that
is in context of the data. Don’t\
forget units!\
2
