---
title: "Untitled"
output: html_document
date: "2024-12-01"
editor_options: 
  markdown: 
    wrap: 72
---

1.) (Use R) Consider the dataset “Homework 6 data.xlsx.” It consists of
5 randomly selected student’s scores on Test 1 and Test 2 in my
introductory statistics course. We want to answer 2 questions:

```{r}
setwd("~/Desktop/Personal_save/Stat_405_Module_14/Module_14_Homework")
HW_6 <- read.csv(file="Homework_6.csv",header=TRUE)
HW_6
```

a.  First, we want to see if there is a difference in the two tests.
    Paired two-tailed t-test

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE)
```

the p-value is less than 0.05, reject the null hypothesis that the means
of Test 1 and Test 2 are equal.

b.  Second, we want to see if there was improvement over the course of
    the semester. H0: Test1 - Test2 \< 0

```{r}
t.test(HW_6$Test.1, HW_6$Test.2, paired = TRUE, alternative = "less")
```

Reject the null hypothesis that the mean difference of Test 1 minus Test
2 is equal to zero. Tentatively conclude that the mean difference of
test 1 minus test 2 is less than zero, and therefore that the the grades
of the second test were better than the first.

**CHECK THIS ONE LATER (Q2)**

2.) (Use R) The data called “plasma” from Anderson et al. (1981)
consists of measurements of plasma concentrations in micromoles/liter
from 10 subjects at times of 8 am, 11am, 2pm, 5 pm, and 8 pm. Analyze
the data in a 1-way ANOVA model choosing time as factor.

```{r}
plasma <- read.csv(file="plasma.csv",header=TRUE)
plasma$time <- factor(plasma$time,levels=c("8am", "11am", "2pm", "5pm", "8pm"), 
                      labels = c("8am", "11am", "2pm", "5pm", "8pm"))

plasma_model <- lm(plasma ~ time, data = plasma)
anova(plasma_model)
```

3.) Two friends play a computer game and each of them repeats the same
level 10 times. The scores obtained are:

```{r}
scores <- read.table(file="scores.txt",header=TRUE)
```
un-paired, two-tailed, t-test 
```{r}

```

a.  Player 2 insists that he is the better player and suggests to
    compare their mean performance. Use a t-test to test whether there
    is a difference between their mean performance (alpha = 0.05).

    ```{r}

    ```

b.  Player 1 insists that he is the better player. He proposes to use
    the Wilcoxon rank-sum test for the comparison. What are the results
    (alpha = 0.05)?
