---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-12-01"
---
1.) (IN R) We have seen the excel file before called Game_Times which shows many variables
associated with 3 days of baseball games in the 2015 season. One reason people think baseball
TV ratings are down is because the games last too long. Suppose we want to find out which
variable has the strongest association with Time of Game to try to make games shorter.
```{r}
setwd("~/Desktop/Personal_save/Stat_405_Module_14/Lab_14.2")
times <- read.csv(file="Game_Times.csv",header=TRUE)
```

a. Examine the dataset graphically and draw conclusions.
```{r}
pairs(times[,c(4,5,6,11)])
pairs(times[,c(7,8,9,11)])
pairs(times[,c(10,12,13,11)])
pairs(times[,c(14,11)])

#variable on the row is on the y-axis, so if we want to see what variable has the greatest effect on time of game, we would look at the bottom row of each of these plots 
```

You have to open it in another the plot in another window but you can see that Time of game and Pitches appears to have the strongest cooralation. 

b. Examine the dataset numerically and draw conclusions.
```{r}
cor(times[,4:14])
```

If you examine the Time.of.Game column you can see that the greatest coloration coefficient is with Pitchers.
