---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-11-25"
---

```{r}
#setwd("C:/Users/jake pc/Desktop/Personal_save/Stat_405_Module_14")
setwd("~/Desktop/Personal_save/Stat_405_Module_14")
test <- read.csv(file="hypothesis.csv",header=TRUE)
```

The following table shows the time for subjects to feel relief from headache pain:

a.) Perform a t-test. Is either product significantly faster than the other at the 0.05 level?

The data is the same as Lab 6.1 question 1 where the samples were independent so the following independent 2 sample t-test is performed.

Two sample, two tailed, independent t-test.

```{r}
test$id <- seq(1:nrow(test))
test
library(reshape2)
long_data <- melt(data = test, id.vars = c("id"),
                   measured.vars = c("Aspirin", "Tylenol"),
                   variable.name = "Brand", value.name = "Time")
long_data
t.test(long_data$Time ~ long_data$Brand)
```

The p-value resulting from the test is greater than 0.05, fail to reject the null hypothesis that the true difference in means between the groups is zero.

b.) Perform a Wilcoxon rank-sum test.
```{r}
wilcox.test(long_data$Time ~ long_data$Brand)
```

The p-value resulting from the test is greater than 0.05, fail to reject the null hypothesis that the true location shift is equal to zero.

