---
title: "Untitled"
output:
  pdf_document: default
  html_document: default
date: "2024-12-01"
editor_options:
  markdown:
    wrap: 72
---

1.) (IN R) Consider the text file called muscle which consists of paired
data of individual’s muscle mass and age. We want to examine this
relationship.

```{r}
setwd("~/Desktop/Personal_save/Stat_405_Module_14/Lab_14.3")
muscle <- read.table(file="muscle.txt",header=TRUE)
```

a.  Begin by looking at a scatterplot and finding the correlation. Does
    the association look linear? How strong is it?

```{r}
muscle
plot(x=muscle$age,y=muscle$mass)
cor(x=muscle$age,y=muscle$mass)
```

The association looks linear, and very strong.

b.  Produce the necessary diagnostic plots to prepare to fit a
    regression model.

fitted values means predicted response values.

-   Residuals vs. Predicted Response

-   Residuals vs. Predictor

-   Studentized Residuals vs. Predictor

-   QQ plot of Residuals

```{r}
library(ggplot2)
model <- lm(muscle$mass ~ muscle$age) 

res <- resid(model)
fit <- fitted(model)

#Residuals vs. Predicted Response 
ggplot(mapping = aes(x = fit, y = res)) +
    geom_point() +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed")

#Residuals vs. Predictor 
ggplot(mapping = aes(x = muscle$age, y = res)) +
    geom_point() +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed")

#Studentized Residuals vs. Predictor observation number 
plot(rstudent(model) ~ seq(1,length(muscle$age),by=1))

#QQ plot of Residuals 
ggplot(mapping = aes(sample = res)) +
    stat_qq() + 
    stat_qq_line()


library(ggfortify)
autoplot(model, which = 1:6, ncol = 2, label.size = 3)
```

c.  Using this information, would you recommend going ahead and fitting
    a simple linear regression model to this data? (Do not fit the model
    yet, just answer yes or no and provide evidence)

    Yes I would recommend fitting a linear model, everything appears
    random with no significant patters. Observation 11 is particularly
    influential in the leverage plot and has a relatively high
    studentised residual but not enough to warrant removal.

<!-- -->

c.  Write a code to fit a simple linear regression model.

    ```{r}
    model <- lm(mass ~ age, data = muscle) 
    summary(model)
    ```

d.  What does the slope mean?

    The meaning of the slope is that as age increases by one year the
    model predicts that muscle mass will decrease by about one unit of
    measurement.

e.  Predict the muscle mass for a 66-year-old person. (Hint: use the
    predict( ) function. Further detail of the predict function for
    linear regression model can be found in the R documentation.
    help(predict.lm))

    ```{r}
    predicted_mass <- predict(model, newdata = data.frame(age = 66))
    predicted_mass
    ```

f.  Calculate a 98% CI for the muscle mass for a 40-year-old. (Hint: use
    the predict( ) function.)

    ```{r}
    prediction_40 <- predict(model, newdata = data.frame(age = 40), 
                                    interval = "confidence", 
                                    level = 0.98)

    prediction_40
    ```
